
\section*{Discussion}

Here, we first showed how tradeoffs, initial conditions, and conflicting/ameliorative coevolution
combine to set the relevant deterministic rules for coexistence in a competitive system.
Tradeoff additivity (i.e., how the cost of investing in multiple axes together 
compares to the sum of investing separately) shapes how fitness maps onto the axis space,
creating stable fitness peak(s) or a neutrally stable ridge.
For multiple peaks or a neutrally stable ridge, initial conditions (i.e., where in the 
axis space species start) determine the peak or ridge location to which each species will evolve.
The extent to which coevolution is conflicting (i.e., investing by one species increases 
competition experience by all others) or ameliorative for each axis affects
how permissible coexistence should be across the axis space.
Thus, tradeoffs and initial conditions dictate where in the axis space evolution will take species, and 
conflicting/ameliorative coevolution determines how well they can 
coexist when they get there.






We also showed how stochasticity could bend and break these deterministic rules.
These effects occurred via transient effects during the critical period
immediately after a new invader arrived to the community.
Effects of stochasticity on new, rare species had a consistently negative
effect on coexistence:
Stochasticity in population dynamics increased the chances of rare species going extinct,
while stochasticity in evolution reduced the species ability for evolutionary rescue.
Stochasticity's effects on the relatively abundant resident species
had varying effects on the chances of coexistence:
Stochasticity in population dynamics could increase or decrease fitness for
the resident species at critical times for an invader, which
could increase or decrease the chances for the invader to survive long enough 
for evolutionary rescue.
Stochasticity in evolution kept phenotypes from remaining at the optimal 
axis values (i.e., $\ddot{v}_{ij} \ne \hat{v}_{ij}$),
which affected resident species in two ways.
First, it could increase or decrease their effects on other species in the community
(see Equation \ref{eq:competition}).
Second, it reduced their fitness and abundance, which had a consistently 
positive effect on coexistence.
This latter effect is why evolution stochasticity had a greater positive effect
on coexistence than population-dynamics stochasticity.
Despite the different mechanisms, both types of stochasticity had similar 
overall effects on coexistence.
When deterministic rules allowed for coexistence among species, 
stochasticity could only have negative effects on it.
On the margins, just outside the bounds of where these rules permitted coexistence, 
stochasticity improved the chances of multi-species coexistence.








% This section is common in AmNat, but not required
% \section*{Conclusions}

